---
title: "Quick Start Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# CovariateSearcher Quick Start

A minimal end-to-end walkthrough: prepare three inputs, run a stepwise covariate
search, read the results, resume if it stops.

For the full reference see the
[online documentation](https://ollegst.github.io/CovariateSearcher/).

---

## What you need

### 1. A base model

A NONMEM control stream whose parameters are named and whose typical values are
written on their own `TV_` lines.

```
$THETA
; Simple names in $THETA - no TV_ prefix here
0.5 ; CL ; L/h ; RATIO
10  ; V  ; L   ; RATIO

$PK
; TV_ prefix HERE, in $PK
TV_CL = THETA(1)
TV_V  = THETA(2)

CL = TV_CL * EXP(ETA(1))     ; normal scale
V  = EXP(TV_V + ETA(2))      ; log scale - both are fine

$OMEGA BLOCK(2)
0.1 ; IIV_CL   ; ; LOG
0.1 ; IIV_CL_V ; ; RATIO
0.1 ; IIV_V    ; ; LOG

$SIGMA
0.1 ; RUV_PROP ; ; LOG
```

`$THETA` line format is `value ; NAME ; units ; RATIO|LOG`, and `$OMEGA BLOCK`
needs **one value per line**:

```
# Wrong                          # Right
$OMEGA BLOCK(2)                  $OMEGA BLOCK(2)
0.1     ; IIV_CL                 0.1 ; IIV_CL   ; ; LOG
0.1 0.1 ; IIV_V                  0.1 ; IIV_CL_V ; ; RATIO
                                 0.1 ; IIV_V    ; ; LOG
```

**The `TV_` line is not cosmetic.** A time-constant covariate goes on the typical
value, and on a log-scale parameter it has to be *added inside* the `EXP()`.
The package checks this at initialization and stops, with a message telling you
how to rewrite, if a parameter you are testing covariates on is written as:

| Rejected | Why |
|---|---|
| `V = EXP(THETA(2) + ETA(2))` | no separate `TV_V` line, so the covariate term would land outside the `EXP` |
| `TV_V = EXP(THETA(2))` with `V = TV_V*EXP(ETA(2))` | log scale disguised as normal - the covariate would be placed multiplicatively on a log value |
| `V = EXP(TV_V + EXP(ETA(2))` | unbalanced parentheses - the transform cannot be detected |

Only parameters that actually receive a covariate are checked.

### 2. An analysis dataset

A standard NONMEM dataset (CSV) containing every covariate you want to test.

### 3. A covariate search table

Rather than writing this by hand, build it from your data - references, levels
and time-dependence are then computed for you:

```{r build_table}
library(CovariateSearcher)

spec <- yspec::ys_load("data/spec/lookup.yml")   # or a plain parsed YAML list
data <- read.csv("data/derived/analysis.csv")

cov_tbl <- build_covariate_reference_table(
  data      = data,
  id        = "ID",
  time      = "TIME",
  Parameter = c("CL",  "CL",     "V"),
  Covariate = c("WT",  "SEX",    "WT"),
  Category  = c("con", "cat",    "con"),
  Formula   = c("power", "linear", "linear"),
  yaml_data = spec
)

write.csv(cov_tbl, "data/derived/covariate_search.csv", row.names = FALSE)
```

The result is one row per parameter-covariate pair:

| PARAMETER | COVARIATE | STATUS | FORMULA | LEVELS | REFERENCE | TIME_DEPENDENT |
|-----------|-----------|--------|---------|--------|-----------|----------------|
| CL        | WT        | con    | power   | NA     | 71.7      | No             |
| CL        | SEX       | cat    | linear  | 0;1    | 1         | No             |
| V         | WT        | con    | linear  | NA     | 71.7      | No             |

| Column | Meaning |
|---|---|
| `PARAMETER` | model parameter the covariate acts on (`CL`, `V`, `KA`, ...) |
| `COVARIATE` | column name in the dataset |
| `STATUS` | **exactly `con` or `cat`** - continuous or categorical |
| `FORMULA` | continuous: `linear`, `power`, `exponential`; categorical: `linear` (one theta per non-reference level) or `power` (covariates whose levels are numeric, e.g. dose); or a custom expression |
| `LEVELS` | `;`-separated **observed level values** for categorical covariates (`0;1`), `NA` for continuous |
| `REFERENCE` | value the effect is normalised to: median (continuous) or most frequent level (categorical) |
| `TIME_DEPENDENT` | `Yes`/`No`, computed from the data; decides whether the effect is placed at individual or population level |
| `INIT` | *optional* initial `$THETA` value - see below |

`cov_to_test` (`beta_WT_CL`, ...) is generated for you; you never write it.

> **`STATUS` must be `con` or `cat`.** Writing `continuous` does **not** raise an
> error - the value is not recognised, so `FORMULA` is then treated as a custom
> expression and you silently get a wrong model. Use `build_covariate_reference_table()`
> and this cannot happen: it rejects anything else up front.

Custom formulas are written as an expression in the reserved names `cov` (the
covariate) and `ref` (its REFERENCE); every other symbol becomes an estimated
THETA:

```{r expr}
Formula = "EMAX*cov/(EC50+cov)"     # two thetas: EMAX, EC50
```

The optional `INIT` column sets the initial estimate of the covariate THETA -
a plain value, a bounded triple, `FIX`, or one entry per theta for a
multi-parameter expression:

```{r init}
INIT = c("0.1", "(0, 0.5, 2)", "0.75 FIX", "EMAX=0.5; EC50=10")
```

The [Covariate Formulas](articles/covariate-formulas.html) article covers all of
this in full: how each form is written into the control stream, categorical
per-level effects, placement on log- and normal-scale parameters, degrees of
freedom, and registering your own formula.

---

## Run a search

### Initialize

```{r initialize}
ss <- initialize_covariate_search(
  base_model_path       = "run1",
  data_file_path        = "data/derived/analysis.csv",
  covariate_search_path = "data/derived/covariate_search.csv",
  models_folder         = "models"
)
```

### Forward then backward (standard SCM)

```{r run_scm}
results <- run_automated_scm_testing(
  search_state   = ss,
  base_model_id  = "run1",
  scm_type       = "standard",   # test every remaining covariate each step
  starting_phase = "forward",
  full_scm       = TRUE          # forward -> backward
)
```

### One phase only

`full_scm = FALSE` runs just the phase you name:

```{r single_phase}
# Forward selection only
results <- run_automated_scm_testing(ss, "run1",
                                     starting_phase = "forward",
                                     full_scm = FALSE)

# Backward elimination only - start from a model that already has covariates
results <- run_automated_scm_testing(ss, "run20",
                                     starting_phase = "backward",
                                     full_scm = FALSE)
```

With `full_scm = TRUE` and `starting_phase = "backward"` you get the full SCM+
sequence: backward, then forward, then backward again.

### Faster: selective SCM

`scm_type = "selective"` carries only the covariates that were significant in
step 1 into later steps, instead of retesting all of them:

```{r selective}
results <- run_automated_scm_testing(ss, "run1", scm_type = "selective",
                                     starting_phase = "forward", full_scm = TRUE)
```

### Thresholds

```{r custom}
results <- run_automated_scm_testing(
  search_state     = ss,
  base_model_id    = "run1",
  forward_p_value  = 0.01,   # stricter entry
  backward_p_value = 0.001,  # stricter retention
  rse_threshold    = 30,     # max RSE (%)
  require_cov_step = TRUE    # require a successful $COVARIANCE step
)
```

---

## What counts as an improvement

A model is kept when **both** hold:

1. **ΔOFV beyond the p-value threshold.** The threshold is a chi-square quantile
   for the covariate's degrees of freedom, so it is *not* a fixed 3.84. df is the
   number of **estimated** (non-`FIX`) thetas the covariate adds:

   | covariate | df | forward threshold at p = 0.05 |
   |---|---|---|
   | continuous, one theta | 1 | 3.84 |
   | categorical, 3 levels | 2 | 5.99 |
   | `EMAX*cov/(EC50+cov)` | 2 | 5.99 |
   | every theta `FIX` | 0 | kept on any ΔOFV > 0 |

2. **Maximum RSE below `rse_threshold`** (default 50%). In backward elimination
   an `NA` RSE is tolerated, for workflows that run without `$COVARIANCE`.

---

## Results

```{r results}
results$final_model        # e.g. "run85"
results$final_covariates   # covariates in the final model
results$excluded_covariates

view_comprehensive_table(results$search_state)   # every model, step by step
```

---

## Save and resume

Checkpoints are written automatically to `<models_folder>/scm_rds/` as
`NN_phase_event.rds`, numbered by step so the highest `NN` is the most recent.
You can also save at any time:

```{r save}
save_search_state(ss, "my_state.rds")   # bare name -> <models_folder>/scm_rds/
```

To resume an interrupted search, point `continue_search()` at the latest
checkpoint. It works out where the search stopped and carries on:

```{r resume}
results <- continue_search(
  checkpoint = "models/scm_rds/07_forward_selection_complete.rds",
  scm_type   = "selective",
  full_scm   = TRUE
)
```

---

## Testing an extra covariate afterwards

To try a covariate that was never in the original table, add it to the live
search state - then the search's history, step numbers and ΔOFV are all kept:

```{r extra}
cov_smk <- build_covariate_reference_table(
  data = data, id = "ID", time = "TIME",
  Parameter = "V", Covariate = "SMK", Category = "cat", Formula = "linear",
  yaml_data = spec
)

ss <- add_covariates_to_search(results$search_state, cov_smk)

res <- add_covariate_to_model(
  ss,
  base_model_id = results$final_model,
  covariate_tag = "beta_SMK_V",
  step_number   = max(ss$search_database$step_number, na.rm = TRUE) + 1,
  phase         = "individual_testing"   # not part of the search itself
)

bbr::submit_model(bbr::read_model(file.path(ss$models_folder, res$model_name)),
                  .bbi_args = list(threads = 12))
```

---

## Common issues

**Base model rejected at initialization.** Read the message: it names the
parameter and the exact rewrite (usually a missing `TV_` line - see the table
above).

**Good models rejected on RSE.** Raise the threshold, or drop the covariance
requirement:

```{r rse_fix}
results <- run_automated_scm_testing(ss, "run1",
                                     rse_threshold = 100,
                                     require_cov_step = FALSE)
```

**A model failed to estimate.** The search creates a retry (`run12` -> `run12001`)
with a different initial estimate for the covariate THETA, and submits it when
`auto_retry = TRUE`.

---

## Where next

```{r help}
?initialize_covariate_search
?run_automated_scm_testing
?build_covariate_reference_table
?continue_search
```

The [online documentation](https://ollegst.github.io/CovariateSearcher/) covers
covariate formulas in depth, the recovery system, forest plots and simulation,
and troubleshooting.
