---
title: "Quick Start Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# CovariateSearcher Quick Start

This is a minimal vignette included in the package installation.

For complete documentation, see the [online documentation](https://ollegst.github.io/CovariateSearcher/).

---

## 📋 What You Need

### 1. Base Model
NONMEM control file with properly formatted parameters:

```
$THETA
; Simple names (NO TV_ prefix in THETA block!)
0.5 ; CL ; L/h ; RATIO
10  ; V  ; L   ; RATIO

$PK
; Use TV_ prefix HERE in PK block
TV_CL = THETA(1)
TV_V  = THETA(2)

CL = TV_CL * EXP(ETA(1))
V  = TV_V  * EXP(ETA(2))

$OMEGA BLOCK(2)
0.1 ; IIV_CL   ; ; LOG
0.1 ; IIV_CL_V ; ; LOG
0.1 ; IIV_V    ; ; LOG

$SIGMA
0.1 ; RUV_PROP ; ; LOG
```

**Parameter Format**: `value ; NAME ; units ; RATIO|LOG`

**Important**: 
- In `$THETA` block: Use `CL`, `V` (simple names) OR `1_CL`, `2_V` (numbered)
- In `$PK/$PRED` block: Use `TV_CL`, `TV_V` (with TV_ prefix)
- **OMEGA BLOCK**: One value per line (not multiple values on same line)
- **Four fields required**: `value ; name ; units ; transform`

**❌ Wrong OMEGA format:**
```
$OMEGA BLOCK(2)
0.1     ; IIV_CL   
0.1 0.1 ; IIV_V   ← Don't put multiple values per line!
```

**✅ Correct OMEGA format:**
```
$OMEGA BLOCK(2)
0.1 ; IIV_CL   ; ; LOG
0.1 ; IIV_CL_V ; ; LOG
0.1 ; IIV_V    ; ; LOG
```

### 2. Analysis Dataset
NONMEM dataset with covariates (CSV format)

### 3. Covariate Search Configuration
CSV file defining which covariates to test:
link : data/derived/covariate_search.csv (optional)

| PARAMETER | COVARIATE | STATUS      | FORMULA | LEVELS | REFERENCE | TIME_DEPENDENT |
|-----------|-----------|-------------|---------|--------|-----------|----------------|
| CL        | WT        | continuous  | power   | NA     | NA        | FALSE          |
| CL        | AGE       | continuous  | linear  | NA     | NA        | FALSE          |
| CL        | SEX       | categorical | linear  | 0;1    | 1         | FALSE          |
| V         | WT        | continuous  | linear  | NA     | NA        | FALSE          |

**Column Descriptions:**
- `PARAMETER`: Model parameter to test covariate on (e.g., CL, V, KA)
- `COVARIATE`: Covariate column name from dataset
- `STATUS`: "continuous" or "categorical"
- `FORMULA`: "linear" or "power" (for continuous covariates)
- `LEVELS`: Number of categories (for categorical), NA otherwise
- `REFERENCE`: Reference category value (for categorical), NA otherwise
- `TIME_DEPENDENT`: TRUE if covariate changes over time, usually FALSE
**Note:** The package automatically generates the `cov_to_test` column (e.g., `beta_WT_CL`) 
during initialization. You don't need to include this column in your CSV file.

---

## 🚀 Basic Usage

### Initialize Search

```{r initialize}
library(CovariateSearcher)

# Initialize a new search
search_state <- initialize_covariate_search(
  base_model_path = "models/run1",
  data_file_path = "data/analysis.csv",
  covariate_search_path = "data/covariates.csv"
)
```

### Run Stepwise Covariate Modeling

```{r run_scm}
# Run standard SCM (Forward → Backward)
results <- run_automated_scm_testing(
  search_state = search_state,
  base_model_id = "run1",
  scm_type = "standard",
  starting_phase = "forward",
  full_scm = TRUE
)
```

### Check Results

```{r results}
# Final model
print(results$final_model)

# Final covariates
print(results$final_covariates)

# View complete database
View(results$search_state$search_database)
```

---

## 📊 Model Selection Criteria

Models are selected if they meet **BOTH** criteria:

1. **ΔOFV > threshold**
   - Forward: ΔOFV > 3.84 (p < 0.05)
   - Backward: ΔOFV > 6.63 (p < 0.01)

2. **RSE < threshold** (default: 50%)

---

## ⚙️ Common Options

### Selective SCM (faster Forward → Backward)

```{r selective}
results <- run_automated_scm_testing(
  search_state = search_state,
  base_model_id = "run1",
  scm_type = "selective",  # Only test significant covariates from step 1
  starting_phase = "forward",
  full_scm = TRUE
)
```

### Full SCM Workflow (Backward → Forward → Backward)

```{r scm_plus}
results <- run_automated_scm_testing(
  search_state = search_state,
  base_model_id = "run1",
  scm_type = "standard",
  starting_phase = "backward",  # Start with backward elimination
  full_scm = TRUE
)
```

### Custom Thresholds

```{r custom}
results <- run_automated_scm_testing(
  search_state = search_state,
  base_model_id = "run1",
  forward_p_value = 0.01,    # More stringent (ΔOFV > 6.63)
  backward_p_value = 0.001,  # Very stringent (ΔOFV > 10.83)
  rse_threshold = 30         # Stricter RSE requirement
)
```

---

## 💾 Save and Resume

### Save Progress

```{r save}
# Save search state
save_search_state(search_state, "scm_state.rds")
```

### Resume Interrupted Search

```{r resume}
# Load saved state
search_state <- load_search_state("scm_state.rds")

# Continue from where you left off
results <- run_automated_scm_testing(
  search_state = search_state,
  base_model_id = "run25",  # Last completed model
  starting_phase = "backward"
)
```

---

## 🔍 Validate Base Model

Check that your base model meets requirements:

```{r validate}
# Validate parameter format
validation <- validate_base_model_parameters(
  base_model_path = "models/run1",
  strict = TRUE
)

# Check validation results
print(validation$valid)
print(validation$issues)
```

---

## ⚠️ Common Issues

### Base model validation fails
**Problem**: Parameter format incorrect

**Solution**: Ensure parameters follow format: `value ; NAME ; units ; RATIO|LOG`

### High RSE rejecting good models
**Problem**: RSE threshold too strict

**Solution**: Increase threshold
```{r rse_fix}
results <- run_automated_scm_testing(
  search_state = search_state,
  base_model_id = "run1",
  rse_threshold = 100  # More lenient
)
```

---

## 📚 Next Steps

Visit the [online documentation](https://ollegst.github.io/CovariateSearcher/) for:

- **User Guide**: Complete reference with all options
- **Complete Workflow**: Detailed step-by-step tutorial
- **Recovery System**: Understanding automatic error handling
- **Troubleshooting**: Solutions to common problems

### Function Documentation

```{r help, eval=FALSE}
?initialize_covariate_search
?run_automated_scm_testing
?validate_base_model_parameters
?save_search_state
?load_search_state
```

---

## 📊 Workflow Summary

```
1. Prepare Files
   ├─ Base model (.ctl)
   ├─ Dataset (.csv)
   └─ Covariate config (.csv)
   
2. Initialize
   └─ initialize_covariate_search()
   
3. Run SCM
   └─ run_automated_scm_testing()
   
4. Check Results
   ├─ results$final_model
   ├─ results$final_covariates
   └─ View(search_state$search_database)
```

---

## ✅ Example Output

After successful SCM run:

```
🎉 AUTOMATED SCM TESTING COMPLETE!
======================================================================
Final model: run85
Final covariates: WT_CL, SEX_V
OFV improvement: -23.45

Models tested: 47
Models completed: 45
Models failed: 2
Models retried: 2

Forward steps: 3
Backward removals: 1
======================================================================
```

---

**That's it!** CovariateSearcher handles the complex workflow automatically.

For detailed examples and advanced usage, see the [complete documentation](https://ollegst.github.io/CovariateSearcher/).
