---
title: "Complete SCM Workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Complete Stepwise Covariate Modeling Workflow

This comprehensive guide walks through a complete SCM analysis from start to finish.

**Note**: This article is NOT included in the package build. You can update it anytime without reinstalling CovariateSearcher!

## Overview

The stepwise covariate modeling (SCM) process consists of:

1. **Forward Selection**: Add covariates that improve model fit
2. **Backward Elimination**: Remove non-significant covariates
3. **Final Model**: Select the best parsimonious model

## Prerequisites

```{r prerequisites}
library(CovariateSearcher)
library(dplyr)
```

## Step 1: Prepare Your Data

### NONMEM Dataset
Your dataset should include:
- `ID`: Subject identifier
- `TIME`: Time points
- `DV`: Dependent variable (concentrations)
- `EVID`: Event ID (0=observation, 1=dose)
- Covariates (WT, AGE, SEX, etc.)

```{r data-example}
# Example dataset structure
head(nonmem_data)
#   ID TIME  DV EVID  WT AGE SEX
#    1  0.0  NA    1  70  45   1
#    1  0.5 5.2    0  70  45   1
#    1  1.0 8.1    0  70  45   1
```

### Covariate Search Configuration

Create a CSV file defining which covariates to test:

```{r covariate-config}
# Example: covariate_search.csv
covariate_config <- tribble(
  ~PARAMETER, ~COVARIATE, ~STATUS, ~FORMULA, ~LEVELS, ~REFERENCE,
  "CL",       "WT",        "test",  "power",  NA,      NA,
  "CL",       "AGE",       "test",  "linear", NA,      NA,
  "CL",       "SEX",       "test",  "cat",    "0,1",   0,
  "V",        "WT",        "test",  "power",  NA,      NA
)

write.csv(covariate_config, "data/derived/covariate_search.csv", row.names = FALSE)
```

## Step 2: Initialize Search

```{r initialize}
search_state <- initialize_new_search(
  base_model_path = "run1",                             # Your base model
  data_file_path = "data/derived/nonmem_data.csv",      # NONMEM dataset
  covariate_search_path = "data/covariate_search.csv",
  models_folder = "models/",                     # Where to save models
  output_folder = "output",                     # Where to save results
  auto_submit = TRUE,                           # Submit models automatically
  auto_retry = TRUE                             # Retry failed models
)
```

## Step 3: Run Forward Selection

```{r forward-selection}
# Standard forward selection
forward_result <- run_stepwise_covariate_modeling(
  search_state = search_state,
  forward_threshold = 3.84,    # p < 0.05 (chi-square with 1 df)
  rse_threshold = 50,          # Max RSE = 50%
  max_steps = 10,              # Maximum forward steps
  submit = TRUE                # Submit to cluster
)

# Check results
print(forward_result$final_model)
print(forward_result$selected_covariates)
```

### Understanding Selection Criteria

Models are selected if they meet BOTH criteria:
- **ΔOFV > 3.84**: Statistically significant improvement (p < 0.05)
- **RSE < 50%**: Parameter estimates are reasonably precise

## Step 4: Backward Elimination

```{r backward-elimination}
backward_result <- run_backward_elimination(
  search_state = search_state,
  backward_threshold = 6.63,   # p < 0.01 (chi-square with 1 df)
  rse_threshold = 50,
  start_model = forward_result$final_model
)

print(backward_result$final_model)
```

## Step 5: Review Results

```{r review}
# Get final model details
final_model_name <- backward_result$final_model

# Extract from database
final_model_row <- search_state$search_database %>%
  filter(model_name == final_model_name)

# Show final covariates
cat("Final covariates:", unlist(final_model_row$tags), "\n")
cat("Final OFV:", final_model_row$ofv, "\n")
cat("OFV improvement from base:", 
    search_state$base_model_ofv - final_model_row$ofv, "\n")
```

## Advanced: Selective Forward Search

If you want to test only covariates that were significant in step 1:

```{r selective}
selective_result <- run_scm_selective_forward(
  search_state = search_state,
  forward_threshold = 3.84,
  rse_threshold = 50
)
```

## Monitoring Progress

While models are running:

```{r monitor}
# Check current status
status_summary <- search_state$search_database %>%
  count(status, phase)
print(status_summary)

# View specific model
model_info <- get_model_info(search_state, "run11")
print(model_info)
```

## Handling Failed Models

CovariateSearcher automatically detects and retries failed models:

```{r recovery}
# Failed models are automatically retried with THETA adjustment
# Check retry models in database
retries <- search_state$search_database %>%
  filter(phase == "retry")

print(retries)
```

## Exporting Results

```{r export}
#read results from final step
search_state <- readRDS("./models/scm_rds/scm_auto_final_complete.rds")
# create table with results
results_table <- create_scm_results_table(search_state)
print_scm_results_table(search_state)
```

## Next Steps

- See [Troubleshooting Guide](troubleshooting.html) for common issues
- See [Recovery System](recovery-system.html) for handling model failures
- See [Case Studies](case-studies.html) for real-world examples

