---
title: "Covariate Formulas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariate Formulas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Covariate Formulas

Everything about how a covariate effect is expressed and written into the
control stream: the built-in forms, custom expressions, initial estimates,
placement, and what each choice costs in degrees of freedom.

The `FORMULA` column of the covariate search table drives all of it. A value is
one of three things:

1. a **built-in name** for a continuous covariate — `linear`, `power`,
   `exponential`
2. a **built-in name** for a categorical covariate — `linear`, `power`
3. anything else — a **user-defined expression** such as `EMAX*cov/(EC50+cov)`

---

## Where the effect goes

Two properties decide how the effect is written, and both are read from your
data and model rather than declared by you.

**Is the covariate time-varying?** If a subject has more than one distinct value
of the covariate, the effect is an individual-level effect and is applied to the
individual parameter line. Otherwise it is a population effect and is applied to
the typical-value (`TV_`) line. `build_covariate_reference_table()` records this
as `TIME_DEPENDENT`.

**Is the parameter log- or normal-scale?** Read from the parameter's own
equation: `PARAM = TV * EXP(ETA)` is normal, `PARAM = EXP(TV + ETA)` is log.

| placement | normal scale | log scale |
|---|---|---|
| population (on `TV_` line) | `*` factor | `+` term |
| individual (on `PARAM =` line) | `*` factor | `*` factor |

A multiplicative factor is only correct on a natural-scale value. On a log-scale
typical value the same effect has to be **added**, because
`EXP(θ + β·LOG(cov/ref))` is what `EXP(θ) * (cov/ref)^β` means. Every built-in
carries both renderings, so this is handled for you.

### What the base model must look like

Because a population covariate is written onto the `TV_` line, a log-scale
parameter needs that line to exist:

```
TV_V1 = THETA(2)
V1    = EXP(TV_V1 + ETA(2))
```

At initialization, `validate_param_transformations()` checks every parameter
that will receive a population covariate and stops on these:

| Rejected | Why |
|---|---|
| `V1 = EXP(THETA(2) + ETA(2))` | no separate `TV_V1` line, so the additive term would land outside the `EXP()` |
| `TV_V1 = EXP(THETA(2))` with `V1 = TV_V1*EXP(ETA(2))` | log scale written in normal-IIV form; the effect would be placed multiplicatively on a log value |
| `V1 = EXP(TV_V1 + EXP(ETA(2))` | unbalanced parentheses — the transformation cannot be detected |

Time-varying covariates are multiplicative on the individual line whatever the
parameterization, so they are not subject to these checks.

---

## Built-in continuous formulas

For covariate `COV` with reference `ref` on THETA number `n`:

| FORMULA | normal scale | log scale |
|---|---|---|
| `power` | `* (COV/ref)**THETA(n)` | `+ THETA(n)*LOG(COV/ref)` |
| `linear` | `* (1 + (COV-ref) * THETA(n))` | `+ LOG(1 + (COV-ref) * THETA(n))` |
| `exponential` | `* EXP(THETA(n) * (COV-ref))` | `+ THETA(n) * (COV-ref)` |

`REFERENCE` is the value at which the effect is 1 (no effect).
`build_covariate_reference_table()` sets it to the median of the baseline values
for a continuous covariate.

To fix an exponent rather than estimate it — allometric scaling, for instance —
use `power` with a fixed `INIT`:

| FORMULA | INIT | effect |
|---|---|---|
| `power` | `1 FIX` | `* (WT/70)**1` |
| `power` | `0.75 FIX` | `* (WT/70)**0.75` |

A fixed theta adds no estimated parameter, which changes how the covariate is
tested — see [Degrees of freedom](#degrees-of-freedom-and-selection).

---

## Categorical covariates

Set `STATUS = "cat"`. `LEVELS` holds the observed level values separated by `;`
(`0;1;2`), and `REFERENCE` is the level the others are compared against — by
default the most frequent one.

### `linear` — one theta per level

Each non-reference level gets its own THETA, written as an `IF/ELSEIF` block
inserted after `$PK`. For `SEX` (levels 1, 2; reference 1) on `CL`:

```
$PK

IF(SEX.EQ.1)THEN
beta_SEX_CL = 1
ELSEIF(SEX.EQ.2)THEN
beta_SEX_CL = 1 + THETA(5)
ENDIF

TV_CL = THETA(1) * beta_SEX_CL
CL    = TV_CL * EXP(ETA(1))
```

On a log-scale parameter the same block is written additively — the reference
level is assigned `0` instead of `1`, non-reference levels get `THETA(n)` rather
than `1 + THETA(n)`, and the term is added to the `TV_` line:

```
IF(SEX.EQ.1)THEN
beta_SEX_CL = 0
ELSEIF(SEX.EQ.2)THEN
beta_SEX_CL = THETA(5)
ENDIF

TV_V1 = THETA(2) + beta_SEX_CL
V1    = EXP(TV_V1 + ETA(2))
```

Non-reference levels are ordered by frequency, most common first, so the
lowest-numbered THETA belongs to the best-informed level.

The `$THETA` entries are named from your lookup file when one is available, and
from the level value otherwise:

```
0.1 ; beta_SEX_CL_FEMALE ;  ; RATIO      # lookup: 2 -> "Female"
0.1 ; beta_SEX_CL_2      ;  ; RATIO      # no lookup entry
```

### `power` — numeric levels

A covariate with discrete but genuinely numeric levels — dose 35/70/125/150 mg,
say — is usually better described by a power relationship on the actual values
than by one theta per level. `STATUS = "cat"` with `FORMULA = "power"` does
that: it renders exactly like the continuous `power` form, with no `IF` block.

---

## User-defined formulas

Any `FORMULA` that is not a built-in name is read as a **single-factor
expression**. Two names are reserved:

- `cov` — the covariate
- `ref` — its `REFERENCE`

Every other symbol becomes an estimated THETA, numbered in order of first
appearance:

| FORMULA | thetas | written into `$PK` |
|---|---|---|
| `1 + slope*log(cov/ref)` | slope | `* (1 + THETA(5) * LOG(WT/70))` |
| `EMAX*cov/(EC50+cov)` | EMAX, EC50 | `* (THETA(5) * WT/(THETA(6) + WT))` |
| `IMAX*cov**hill/(IC50**hill+cov**hill)` | IMAX, hill, IC50 | `* (THETA(5) * WT**THETA(6)/(THETA(7)**THETA(6) + WT**THETA(6)))` |

Allowed in an expression: `+`, `-`, `*`, `/`, `^` (or `**`), and the functions
`exp`, `log`, `log10`, `sqrt`. Conditionals (`if`, `ifelse`) are rejected, which
keeps every expression a single factor — per-level behaviour is what
`cat`/`linear` is for.

An expression must reference `cov`, and must contain at least one symbol that is
not `cov`/`ref`; otherwise there is nothing to estimate.

```{r expr_example}
cov_tbl <- build_covariate_reference_table(
  data      = data,
  id        = "ID",
  time      = "TIME",
  Parameter = "CL",
  Covariate = "AGE",
  Category  = "con",
  Formula   = "EMAX*cov/(EC50+cov)",
  yaml_data = spec,
  INIT      = "EMAX=0.5; EC50=10"
)
```

Each theta gets its own `$THETA` line, named after the symbol:

```
0.5 ; beta_AGE_CL_EMAX ;  ; RATIO
10  ; beta_AGE_CL_EC50 ;  ; RATIO
```

### Expressions and log-scale parameters

The package chooses the **operator** by the parameter's scale, and inserts your
expression **exactly as written**. It never transforms it.

| parameter | joined with |
|---|---|
| `CL = TV_CL * EXP(ETA(1))` (normal scale) | `*` |
| `V = EXP(TV_V + ETA(2))` (log scale) | `+` |

So on a log-scale parameter the expression must already be on the log scale —
writing it that way is your responsibility. Use `log()` in the expression and it
is rendered as NONMEM's `LOG()`:

```{r log_expr}
Formula = "log(EMAX*cov/(EC50+cov))"      # on a log-scale parameter
```

which produces

```
TV_V = THETA(2) + (LOG(THETA(3) * AGE/(THETA(4) + AGE)))
V    = EXP(TV_V + ETA(2))
```

The same expression **without** `log()` is still inserted with `+`, giving
`TV_V = THETA(2) + (THETA(3) * AGE/(THETA(4) + AGE))` — a different model, and
almost certainly not what you meant. Whenever an expression goes onto a
log-scale parameter the log records:

```
WARNING: user expression added to log-scale parameter 'V' with '+'; the
expression is inserted as written, so it must already be on the log scale
(e.g. LOG(...)).
```

The built-in forms need no such care: each carries its own additive rendering
and switches to it automatically.

---

## The INIT column

`INIT` sets the initial estimate written for the covariate's THETA. It is copied
into the `$THETA` record as given, so anything NONMEM accepts works:

| INIT | `$THETA` line | meaning |
|---|---|---|
| *(blank)* | `0.1 ; beta_WT_CL` | formula default |
| `0.5` | `0.5 ; beta_WT_CL` | plain initial estimate |
| `(0, 0.5, 2)` | `(0, 0.5, 2) ; beta_WT_CL` | lower bound, initial, upper bound |
| `0.75 FIX` | `0.75 FIX ; beta_WT_CL` | fixed, not estimated |
| `EMAX=0.5; EC50=10` | one line per theta | multi-theta expression |

For a multi-theta expression the entries are matched **by name**, so order does
not matter and any theta you leave out defaults to `0.1`:

```{r init_named}
INIT = "EC50=10; EMAX=0.5"     # same as "EMAX=0.5; EC50=10"
INIT = "EC50=10"               # EMAX falls back to 0.1
```

For a categorical covariate a single `INIT` value is shared by every level's
theta.

**`INIT` must match the number of THETAs the formula declares**, and the table
check enforces it:

| FORMULA | valid INIT | rejected |
|---|---|---|
| `power`, `linear`, `exponential`, `cat.power`, `cat.linear` | one estimate: `0.1`, `(0, 0.5, 2)`, `0.75 FIX` | anything named or multi-entry, e.g. `EMAX=0.5; EC50=10` |
| `EMAX*cov/(EC50+cov)` | one named entry per theta: `EMAX=0.5; EC50=10` | a bare value, or a name the formula does not declare |

```
Row 1: FORMULA 'power' for covariate 'AGE' on parameter 'CL' has a single THETA,
so INIT takes one initial estimate, but got a named/multi-entry spec:
'EMAX=0.5; EC50=10'.
  Use e.g. "0.1", "(0, 0.5, 2)" or "0.75 FIX".
```

---

## Degrees of freedom and selection

A covariate is kept when its ΔOFV beats a chi-square threshold. The threshold
depends on how many parameters the covariate actually adds, so it is not a fixed
3.84.

**df = the number of estimated (non-`FIX`) thetas the covariate adds.**

| covariate | df | forward threshold at p = 0.05 |
|---|---|---|
| continuous, one theta | 1 | 3.84 |
| categorical, 3 levels | 2 | 5.99 |
| `EMAX*cov/(EC50+cov)` | 2 | 5.99 |
| `power` with `INIT = "0.75 FIX"` | 0 | see below |

```{r df}
calculate_covariate_df("WT", search_state$covariate_search)
pvalue_to_threshold(0.05, df = 2)
#> 5.991465
```

A covariate whose thetas are all `FIX` estimates nothing, so a likelihood-ratio
test is meaningless. Its df is 0, the threshold is 0, and it is kept whenever
the model improves at all — any ΔOFV > 0. This is what makes fixed allometric
scaling testable: it either helps the fit or it does not.

---

## Reading the results

THETA names encode what the covariate is and where it acts:

| `$THETA` name | meaning |
|---|---|
| `beta_WT_CL` | single-theta effect of WT on CL |
| `beta_SEX_CL_FEMALE` | categorical level "Female" of SEX on CL |
| `beta_AGE_CL_EMAX` | the `EMAX` theta of a multi-theta expression |

The same names appear as `cov_to_test` in the search table (`beta_WT_CL`) and as
bbr tags on the models, which is how a model's covariates are recovered from
disk.

---

## See also

- [Quick Start](../quick-start.html) — the minimal path through a search
- `?build_covariate_reference_table` — building the table from your data
- `?validate_covariate_search_table` — the checks applied to the table
- `?calculate_covariate_df`, `?pvalue_to_threshold`
