---
title: "Forest Plots and Simulations"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Forest Plots and Simulations

Covariate effects visualised two ways, from the same starting point:

```
sample_individual_thetas -> build_scenario_parameters -+-> plot_parameter_forests
                                                       |
                                                       +-> simulate_scenario_profiles
                                                              -> metrics -> plot_exposure_forest
```

**Parameter forests** need no simulation - the covariate effect on `CL`, `V2`, …
comes straight out of the model. **Exposure forests** carry those parameters
through an mrgsolve model to `AUC`/`Cmax`/`Cmin`.

Both vary one covariate at a time around a typical subject, using parameter
uncertainty. For exposure in the *real* population, split by covariate group,
see [Covariate Boxplots](covariate-boxplots.html).

---

## Requirements

| what | why |
|---|---|
| `.ext` **and** `.cov` (or `.cor`) next to the model | parameter uncertainty is drawn MVN from the covariance matrix; without it each THETA is drawn independently from its `.ext` standard error, with a warning |
| `data` **pre-filtered to the modelling population** | the model's `$DATA IGNORE`/`ACCEPT` is **not** applied - references and levels are computed from the rows you pass (the function warns) |
| a yspec spec | axis labels, units and decoded categorical levels |
| `mrgsolve` (exposure forests only) | in `Suggests` - install it yourself |
| `devEMF` (for `.emf` output) | in `Suggests` |

### The simulation model

Every parameter draw is simulated as its own subject (`idata`) under a common
dose (`events`), so the mrgsolve model has to be written for that:

1. **`$PARAM` declares `THETA1..THETAn`.** The numbers written there are only
   placeholders. For every simulated subject they are replaced by one row of the
   parameter table, which holds the **final individual parameter values** -
   natural scale, covariate effect already applied, one row per uncertainty
   draw. The names must match the table's columns.
2. **Assign the structural parameters straight from them** -
   `double CL = THETA3;`. **No covariate terms and no `EXP()`**: both are
   already in the incoming values, and repeating either double-counts.
3. **No between-subject variability** - no `$OMEGA`, no `ETA`. The spread you
   want is the parameter uncertainty already in the draws; adding random effects
   layers IIV on top of it.
4. **`$CAPTURE` the concentration** you will derive metrics from.
5. `$SIGMA`/`EPS` only if you also want a noisy `DV`; exposure metrics are
   normally taken from the noise-free concentration.

```c
$PARAM
THETA1 = 2.8,
THETA2 = 160,
THETA3 = 32,
THETA4 = 1,
THETA5 = 0.5

$CMT  @annotated
DEPOT    : Dose compartment
TRANSIT  : Transit compartment 1
TRANSIT2 : Transit compartment 2
CENTRAL  : Central compartment

$PLUGIN autodec, nm-vars

$SIGMA 1

$MAIN
double KA   = THETA1;
double V1   = THETA2;
double CL   = THETA3;
double ADD  = THETA4;
double PROP = THETA5;

double K12 = KA;
double K23 = KA;
double K34 = KA;
double K40 = CL/V1;
double S4  = V1/1000.0;

$ODE
dxdt_DEPOT    = -K12*DEPOT;
dxdt_TRANSIT  =  K12*DEPOT    - K23*TRANSIT;
dxdt_TRANSIT2 =  K23*TRANSIT  - K34*TRANSIT2;
dxdt_CENTRAL  =  K34*TRANSIT2 - K40*CENTRAL;

$TABLE
double CP    = CENTRAL/S4;
double IPRED = CP;
double W     = sqrt(ADD*ADD + pow(PROP*IPRED,2));
double DV    = IPRED + W*EPS(1);

$CAPTURE @annotated
CP    : Plasma concentration (ng/mL)
IPRED : Individual prediction
DV    : Simulated observation
CL    : Clearance (L/h)
V1    : Central volume (L)
```

Note `THETA4`/`THETA5` here are the residual-error thetas: they come through the
same parameter table, so the `$THETA` order in the NONMEM model and the
`THETA<n>` names in `$PARAM` must line up.

---

## Inputs

```{r read_data}
library(CovariateSearcher)
library(yspec)     # ys_load, pull_meta
library(yaml)      # read_yaml
library(here)      # here()
library(dplyr)

## The covariate search table the model was built from - it supplies the
## covariates to vary, their REFERENCE, and the LEVELS of categorical ones
covariate_search <- read.csv("./data/derived/covariate_search.csv")

## Two spec files, used for different things:
##   lookup.yml    - covariate names, units and level decodes -> scenario labels
##   pk-extend.yml - PK parameter labels and units            -> param_info
lookup  <- ys_load(here("data", "spec", "lookup.yml"))
spec_pk <- read_yaml(here("data", "spec", "pk-extend.yml"))

## Analysis dataset, reduced to the modelling population and one row per
## subject. $DATA IGNORE/ACCEPT is NOT applied for you, so filter here.
dat <- read.csv("./data/derived/analysis.csv", stringsAsFactors = FALSE) %>%
  filter(EXFLG == 0) %>%
  group_by(ID) %>%
  slice(1)

## Decode categorical covariates to their labels
flags <- pull_meta(lookup, "flags")
dat   <- decode_dataset(dat, lookup, flags$diagCatCov)

model_folder <- "models"
runno        <- "run16"

set.seed(12345)
```

Decoding is safe here: continuous covariates are the only thing read from
`dat` (their quantiles), while categorical levels and references come from
`covariate_search`, so they stay numeric for the model's `IF(SEX.EQ.1)` blocks.

`sample_individual_thetas()` seeds itself through its own `seed` argument; the
global `set.seed()` covers anything else in the script.

---

## Parameter forests

```{r param_forest}
# 1. Parameter uncertainty: MVN draws from the model's .ext + covariance matrix
thetas <- sample_individual_thetas(
  model         = runno,
  models_folder = model_folder,
  Nsamples      = 10000,         # bump to 1e5 for the final figure
  seed          = 1234           # fixed by default, so the draws reproduce
)

attr(thetas, "sampling_method")  # "mvn", or "independent" if no .cov/.cor

# 2. Per-scenario structural parameters, covariate effects baked in.
#    `data` must already be filtered to the modelling population.
params <- build_scenario_parameters(
  model               = runno,
  covariate_search    = covariate_search,   # object or path (.csv/.rds)
  thetas              = thetas,
  data                = dat,                # object or path; already filtered
  spec                = lookup,             # units + decoded levels on the labels
  percentiles         = c(0.1, 0.9),        # any number of quantiles, not just two:
                                            # c(0.05, 0.25, 0.75, 0.95) gives 4 scenarios
                                            # per continuous covariate
  id_col              = "ID",
  scenario_table_path = paste0("results/models/", runno, "/forestplots/rds/")
)

# 3. One forest per structural parameter, saved + combined PDF
plot_parameter_forests(
  params,
  model              = runno,
  output_plot_folder = paste0("results/models/", runno, "/forestplots/plots/"),
  output_rds_folder  = paste0("results/models/", runno, "/forestplots/rds/"),
  output_format      = c("emf", "png"),
  param_info         = spec_pk,
  combined_pdf       = TRUE,
  typical_subject    = TRUE,     # subtitle describing the reference subject;
                                 # FALSE removes it, or pass your own string

  width              = 7,
  height             = 6
)
```

`build_scenario_parameters()` returns a **named list** - one data frame of
per-sample parameters per scenario, the first being the typical subject (all
covariates at `REFERENCE`).

| argument | default | effect |
|---|---|---|
| `percentiles` | `c(0.05, 0.95)` | quantiles of each **continuous** covariate to build scenarios from. Any length: `c(0.05, 0.25, 0.75, 0.95)` gives four scenarios per covariate. Categorical covariates instead get one scenario per non-reference level |
| `models_folder` | `"models"` | where the model and its `.ext`/`.cov` live |
| `id_col` | `"ID"` | de-duplicates `data` to one row per subject before computing the quantiles |
| `lookup`, `spec` | `NULL` | covariate names, units and decoded levels used in the scenario labels |
| `wrap_width` | `30` | break a long scenario label onto a second line; `Inf`/`NULL` disables |
| `scenario_table_path` | `scenario_table_<model>.rds` | where the scenario table is saved. A **directory** gets `scenario_table_<model>.rds` written into it |

Note `percentiles` means something different in the plot functions, where it is
the interval used in the **summary tables** (`c(0.05, 0.95)` → median `[5th,
95th]`).

`plot_parameter_forests()` options:

| argument | default | effect |
|---|---|---|
| `parameters` | `NULL` | restrict to named parameters; `NULL` plots all |
| `output_plot_folder` / `output_rds_folder` | `"results/figure"` / `NULL` | plots and summary-table RDS can go to different folders |
| `output_format` | `c("emf","png")` | one file per format |
| `combined_pdf` | `TRUE` | also write every forest into one multi-page PDF |
| `outer_range` | `NULL` | second, wider shaded band (ratio scale) |
| `typical_subject` | `TRUE` | subtitle describing the reference subject, taken from the attribute `build_scenario_parameters()` attaches. `FALSE`/`NULL` shows no subtitle; a character string is used verbatim |
| `param_info` | `NULL` | labels/units - a named `list(label=, unit=)` or a spec. Supplying it also writes the summary tables |
| `scenario`, `lookup` | `NULL` | not needed here: `build_scenario_parameters()` attaches the scenario table and lookup as attributes, and they are read back automatically |
| `reference` | `"Typical subject"` | scenario used as the ratio denominator |
| `percentiles` | `c(0.05, 0.95)` | interval shown in the summary tables |
| `models_folder` | `"models"` | used to map THETA positions to parameter names |
| `verbose` | `TRUE` | print each parameter as it is plotted |
| `width`, `height` | `6`, `6` | inches |

Parameters with no covariate acting on them are skipped, and each forest shows
only the scenarios that actually move that parameter.

---

## Exposure forests

Add a simulation step. The model is compiled once, by you:

```{r exposure}
library(mrgsolve)

mod  <- mread("models/sim_model.cpp")                     # structural only
dose <- ev(amt = 300, cmt = 1, ii = 24, addl = 6, ss = 1)

profiles <- simulate_scenario_profiles(
  param_sets = params,
  mod        = mod,
  dose       = dose,
  start      = 0,
  end        = 24,       # one dosing interval for steady-state AUC
  delta      = 0.1,
  verbose    = TRUE      # per-scenario progress line
)
```

`profiles` is one row per (scenario, sample, time). Reduce it to per-subject
metrics yourself - the package does not impose a definition:

```{r metrics}
metrics <- profiles |>
  dplyr::group_by(Scenario, ID) |>
  dplyr::summarise(
    AUC  = sum(diff(time) * (head(CP, -1) + tail(CP, -1)) / 2),
    Cmax = max(CP),
    Cmin = min(CP),
    .groups = "drop"
  )

saveRDS(metrics, paste0("results/models/", runno, "/exposure_metrics.rds"))
```

Then one forest per metric:

```{r exposure_forest}
for (m in c("AUC", "Cmax", "Cmin")) {
  plot_exposure_forest(
    metrics,
    metric      = m,
    ss          = TRUE,   # steady state: "ss" is appended to the metric name in
                          # the auto title AND to the saved file name, so the
                          # ss and non-ss versions never overwrite each other.
                          # Applied to whatever `metric` is, so a parameter
                          # would read "CLss" - use ss = FALSE there.
    scenario    = attr(params, "scenario_table"),  # the scenario TABLE, not the
                                             # parameter list: sets the axis order,
                                             # supplies the typical-subject
                                             # subtitle, and is required for the
                                             # summary tables. The saved
                                             # scenario_table_<model>.rds works too
    metric_info = spec_pk,                   # labels/units -> also writes the tables
    model       = runno,
    reference   = "Typical subject",   # scenario used as the ratio denominator;
                                       # it stays as a row - `typical_subject`
                                       # only controls the subtitle

    filename    = paste0("results/models/", runno, "/forestplots/plots/", m, "_forest"),
                          # no need to write "ss" yourself - with ss = TRUE the
                          # saved file becomes AUC_forest_ss.emf / .png
    output_format = c("emf", "png"),
    width = 7, height = 6
  )
}
```

| argument | default | effect |
|---|---|---|
| `metric` | `"AUC"` | any numeric column of `data`, not only exposure metrics |
| `ss` | `TRUE` | steady state. Appends "ss" to the metric name in the auto-generated title (`"Covariate effects on AUCss"`) **and** to the saved file stem (`AUC_forest` → `AUC_forest_ss.emf`), so the two versions cannot overwrite each other. Applied to any `metric`, so set `FALSE` for parameters. The title part is ignored when `title` is given |
| `ClinicalRelevanceLow` | `0.8` | lower bound of the inner shaded band (ratio scale) |
| `ClinicalRelevanceHigh` | `1.25` | upper bound of the inner shaded band |
| `outer_range` | `c(0.5, 2)` | outer band; `NULL` removes it |
| `reference` | `"Typical subject"` | denominator scenario, highlighted |
| `scenario` | `NULL` | the scenario **table** - a data frame, or a path to the saved `scenario_table_<model>.rds`. Sets the axis order, supplies the typical-subject subtitle, and is required before any summary table is written. A plain character vector gives the order only. Needed here because `group_by()`/`summarise()` dropped both |
| `x_lim` | `NULL` | fix the ratio axis |
| `x_breaks` | `NULL` | tick positions on the ratio axis, e.g. `c(0.5, 0.8, 1, 1.15, 1.25)`; `NULL` uses ggplot2's default breaks |
| `fontsize` | `9` | base font size |
| `title` | `NULL` | built from the metric when `NULL` |
| `filename` | `NULL` | no file written when `NULL`; `width`/`height` are required when set |

Because `metric` accepts any numeric column, a parameter forest can also be
drawn this way:

```{r stacked}
long <- stack_scenario_parameters(params, model = runno)
plot_exposure_forest(long, metric = "CL", ss = FALSE, width = 6, height = 6)
```

---

## Summary tables

Both forest functions can write **two tables per run - absolute and relative**.
They are produced only when you pass the labels (`param_info` / `metric_info`);
without them you get plots only.

|  | parameter forests | exposure forests |
|---|---|---|
| trigger | `param_info` | `metric_info` + `filename` + `scenario` |
| files | `<model>-parameter-table-absolute.rds`<br>`<model>-parameter-table-relative.rds` | `<model>-exposure-table-absolute.rds`<br>`<model>-exposure-table-relative.rds` |
| written to | `output_rds_folder` | the plot's folder (`dirname(filename)`) |
| contents | a **named list, one table per parameter** | one table covering **every numeric column** of `data` except `ID` |

Both tables have the same shape: one row per scenario, the covariate columns
that vary across those scenarios (decoded, with units in the header), then one
column per quantity headed `label [unit]`.

The cells differ:

| table | cell | meaning |
|---|---|---|
| absolute | `median`<br>`[lo, hi]`<br>`geomean`<br>`(geoCV%)` | the distribution of the uncertainty draws for that scenario. `lo`/`hi` are `percentiles`; the geometric mean and geometric CV% use the positive values only |
| relative | `ratio [lo, hi]` | median, `lo` and `hi` each divided by the **reference scenario's median** for that quantity - so the reference row reads `1 [...]` |

Everything is rounded to 2 dp and stored as formatted strings, ready to drop
into a report table.

```{r tables}
rds_dir <- paste0("results/models/", runno, "/forestplots/rds/")

## absolute: median / [lo, hi] / geomean / (geoCV%) per scenario
abs_tbl <- readRDS(file.path(rds_dir, paste0(runno, "-parameter-table-absolute.rds")))

## relative: ratio [lo, hi] versus the reference scenario's median
rel_tbl <- readRDS(file.path(rds_dir, paste0(runno, "-parameter-table-relative.rds")))

names(abs_tbl)     # one entry per parameter
abs_tbl$CL
rel_tbl$CL         # same rows and covariate columns, ratios instead of values
```

The exposure pair is read the same way, from the plot folder:

```{r tables_exposure}
plot_dir <- paste0("results/models/", runno, "/forestplots/plots/")

abs_exp <- readRDS(file.path(plot_dir, paste0(runno, "-exposure-table-absolute.rds")))
rel_exp <- readRDS(file.path(plot_dir, paste0(runno, "-exposure-table-relative.rds")))
```

Every quantity must have a label, so the metric/parameter column names have to
match the spec keys - a missing one stops the call rather than silently
producing an unlabelled column.

---

## Function reference

| function | in | out |
|---|---|---|
| `sample_individual_thetas()` | model name | `data.frame(ID, THETA1..n)`, estimation scale |
| `create_covariate_table()` | search table + data | scenario table (one row per scenario) |
| `apply_covariate_model()` | thetas + one scenario | natural-scale structural parameters |
| `build_scenario_parameters()` | the three above | named list of per-scenario parameter sets |
| `stack_scenario_parameters()` | that list | one long data frame with a `Scenario` column |
| `simulate_scenario_profiles()` | parameter sets + mrgsolve model | stacked concentration-time profiles |
| `plot_parameter_forests()` | parameter sets | one forest per parameter (+ tables) |
| `plot_exposure_forest()` | metrics table | one forest per metric (+ tables) |

---

## See also

- [Covariate Boxplots](covariate-boxplots.html) - exposure in the real population, by covariate group
- [Complete Workflow](complete-workflow.html) - producing the model these plots describe
- [Covariate Formulas](covariate-formulas.html) - how the effects being plotted are parameterised
