---
title: "Generating Parameter Tables with model_report()"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Generating Formatted Parameter Tables

The `model_report()` function creates professional parameter tables from NONMEM model outputs. It formats estimates, RSE values, shrinkage, and includes model-level metrics like OFV and Conditional number in a publication-ready table.

## 📋 Overview

**Purpose:** Generate formatted parameter comparison tables for one or more models

**Key Features:**
- ✅ Formatted parameter names and labels from YAML specs
- ✅ Automatic grouping (typical parameters, IIV, residual error, covariates)
- ✅ Side-by-side comparison of multiple models
- ✅ Handles failed models gracefully
- ✅ Publication-ready flextable output

---

## 🚀 Basic Usage

### Step 1: Create `spec_pk` as an R Object

The `model_report()` function expects `spec_pk` to be an in-memory R object (list-like), not a file path string.

```{r load_spec}
library(CovariateSearcher)
library(here)
library(yaml)

# Load parameter specification into an R object
spec_pk <- yaml::read_yaml(here::here("data", "spec", "pk-extend.yml"))

# Then pass the object to model_report(spec_pk = spec_pk)
```

**What's in pk-extend.yml?**
This file defines how parameters should be displayed:

```yaml
CL:
  short: CL/F
  label: Apparent clearance
  unit: L/h
V:
  short: Vc/F
  label: Apparent central volume
  unit: L
IIV_CL:
  short: CL CV%
  label: Inter-individual variability on clearance
```

See the [YAML Specification Files](yaml-specification-files.html) article for complete details.

---

### Step 2: Generate Report for Single Model

```{r single_model}
# Generate table for one model
pk_res <- model_report(
  model_names = "run28",
  spec_pk = spec_pk
)

# Display table
pk_res
```

**Output Example:**

| parameter_names         | label                      | Estimate | RSE (%) | Shrinkage (%) |
|-------------------------|----------------------------|----------|---------|---------------|
| **Typical parameters**                                                                    |
| Ka (1/hr)               | Absorption rate constant   | 4.91     | 20.35   |               |
| CL/F (L/h)              | Apparent clearance         | 121.85   | 4.76    |               |
| Vc/F (L)                | Apparent central volume    | 479.42   | 6.29    |               |
| **Inter-individual variability**                                                          |    
| KA CV%                  | IIV on absorption          | 615.49   | 49.21   | 10.16         |
| CL/F CV%                | IIV on clearance           | 52.22    | 4.48    | 2.82          |
| **Parameter-Covariate relationships**                                                     |
| AGE~CL/F                | Effect of AGE on clearance | -0.06    | 39.26   |               |
| **Residual variability**                                                                  |
| prop error              | Proportional error         | 34.39    | 0.51    | 12.10         |

---

### Step 3: Generate Report for Multiple Models

Compare parameters across multiple models:

```{r multiple_models}
# Compare 3 models side-by-side
pk_res <- model_report(
  model_names = c("run10", "run28", "run45"),
  spec_pk = spec_pk
)

pk_res
```

**Output shows:**
- Side-by-side comparison
- Each model in separate column
- Format: `Estimate (RSE%) [Shrinkage%]`

---

## 📊 Function Parameters

```{r function_signature}
model_report(
  model_names,               # Character vector of model names
  shrinkage = "etasd",      # Type of shrinkage ("etasd", "etavr", "ebvsd", "ebvvr")
  models_folder = "models", # Path to models folder
  spec_pk = NULL,            # Optional R object with parameter metadata
  lookup = NULL              # Optional covariate lookup spec for decoding categorical levels
)
```

### Parameters Explained

| Parameter       | Description                          | Default    | Required    |
|-----------------|--------------------------------------|------------|-------------|
| `model_names`   | Model name(s) without file extension | -          | ✅ Yes |    |
| `shrinkage`     | Type of shrinkage to report          | `"etasd"`  | No          |
| `models_folder` | Directory containing models          | `"models"` | No          |
| `spec_pk`       | R object (list-like) for formatting metadata | `NULL` | Recommended |
| `lookup`        | R object (list-like) with covariate `values`/`decode` for decoding categorical levels | `NULL` | Optional |

### Shrinkage Options

- `"etasd"` - ETA-based SD (default, most common)
- `"etavr"` - ETA-based variance
- `"ebvsd"` - EBV-based SD
- `"ebvvr"` - EBV-based variance

---

## 🏷️ Decoding Categorical Covariate Levels

For categorical covariates, a separate THETA is estimated per non-reference level
(e.g. `beta_SEXN_V_2`). By default the table labels these by their raw level value:

| parameter_names    | label                                          |
|--------------------|------------------------------------------------|
| SEXN level 2~Vc/F  | Effect of SEXN level 2 on Apparent central volume |

Pass a `lookup` object (the same `lookup.yaml`-shaped spec used elsewhere, where each
covariate entry carries `values` and `decode`) to replace the generic level number with
the decoded category name:

```{r decode_levels}
# lookup.yaml contains, e.g.:
#   SEXN:
#     values: [1, 2]
#     decode: ["Male", "Female"]
lookup <- yaml::read_yaml(here::here("data", "spec", "lookup.yaml"))

pk_res <- model_report(
  model_names = "run28",
  spec_pk = spec_pk,
  lookup = lookup
)
```

| parameter_names    | label                                          |
|--------------------|------------------------------------------------|
| SEXN Female~Vc/F   | Effect of Female (SEXN) on Apparent central volume |

When `lookup` is `NULL`, or a covariate/level is not found in it, the original
"level N" label is used, so existing reports are unaffected.

---

## 🎯 Complete Example

```{r complete_example}
library(CovariateSearcher)
library(here)
library(yaml)

# 1. Load YAML specification into an R object
spec_pk <- yaml::read_yaml(here::here("data", "spec", "pk-extend.yml"))

# 2. Generate report for final models
final_models <- c("run85", "run90", "run95")

pk_res <- model_report(
  model_names = final_models,
  spec_pk = spec_pk
)

# 3. Display table
pk_res

# 4. Save to Word document (optional)
flextable::save_as_docx(
  pk_res, 
  path = "output/parameter_table.docx"
)

# 5. Save to PowerPoint (optional)
flextable::save_as_pptx(
  pk_res,
  path = "output/parameter_table.pptx"
)
```

---

## 🛡️ Error Handling

The function gracefully handles models that fail to load or process.

### Example: Some Models Fail

```{r error_handling}
# run1 doesn't exist, but others do
pk_res <- model_report(
  model_names = c("run1", "run28", "run45"),
  spec_pk = spec_pk
)
```

**Output:**
```
📊 Generating report for 3 model(s)...
  Processing run1... ✗
    ⚠️  Error: File does not exist: 'models/run1.yaml'
  Processing run28... ✓
  Processing run45... ✓

⚠️  WARNING: The following models failed and were excluded:
  • run1

✅ Successfully processed 2/3 model(s)

[Table shows run28 and run45 only]
Note: 1 model(s) excluded due to errors: run1
```

**Key Points:**
- ✅ Failed models are automatically skipped
- ✅ Clear indication which models failed
- ✅ Detailed error message for each failure
- ✅ Report generated for successful models only
- ✅ Footer note shows excluded models

---

## 📝 Parameter Formatting

### Without spec_pk (Basic)

```{r without_spec}
# No YAML specification
pk_res <- model_report(
  model_names = "run28",
  spec_pk = NULL  # or just omit this parameter
)
```

**Output shows:**
- Parameter names as they appear in NONMEM (e.g., `THETA1`, `OMEGA(1,1)`)
- No formatted labels
- No units

### With spec_pk (Formatted)

```{r with_spec}
# With YAML specification
spec_pk <- yaml::read_yaml(here::here("data", "spec", "pk-extend.yml"))

pk_res <- model_report(
  model_names = "run28",
  spec_pk = spec_pk
)
```

**Output shows:**
- Formatted names: `CL/F (L/h)` instead of `THETA1`
- Descriptive labels: "Apparent clearance"
- Proper units from YAML file
- Covariate effects formatted nicely

**Recommendation:** Always use `spec_pk` for professional output!

---

## 🎨 Customizing Output

### Modify Table Appearance

The output is a `flextable` object, so you can customize it:

```{r customize}
pk_res <- model_report(
  model_names = c("run28", "run45"),
  spec_pk = spec_pk
)

# Customize the table
pk_res <- pk_res %>%
  flextable::fontsize(size = 10, part = "all") %>%
  flextable::color(color = "darkblue", part = "header") %>%
  flextable::add_header_lines(
    values = "Population PK Parameter Estimates"
  )

pk_res
```

### Export Options

```{r export}
# To Word
flextable::save_as_docx(
  pk_res, 
  path = "tables/parameters.docx"
)

# To PowerPoint
flextable::save_as_pptx(
  pk_res,
  path = "tables/parameters.pptx"
)

# To HTML
flextable::save_as_html(
  pk_res,
  path = "tables/parameters.html"
)

# To image
flextable::save_as_image(
  pk_res,
  path = "tables/parameters.png"
)
```

---

## 🔍 Understanding the Output

### Table Sections

The table is automatically organized into sections:

1. **Typical parameters** - Population (fixed effects) parameters
2. **Inter-individual variability** - Random effects (ETAs/OMEGAs)
3. **Correlation of random effects** - Off-diagonal OMEGAs
4. **Parameter-Covariate relationships** - Covariate effects (betas)
5. **Residual variability** - Residual error (SIGMAs)

### Column Meanings

**Single Model:**
- `parameter_names`: Formatted parameter name
- `label`: Full description
- `Estimate`: Parameter value
- `RSE (%)`: Relative standard error
- `Shrinkage (%)`: ETA shrinkage (for random effects)
- `OFV` row: Objective function value
- `Conditional number` row: Matrix-conditioning metric from `.cor`/`.cov` outputs

**Multiple Models:**
- `parameter_names`: Formatted parameter name
- `label`: Full description
- `Model X`: Combined as `Estimate (RSE%) [Shrinkage%]`

### Special Notations

- `FIX` - Parameter was fixed in estimation
- Empty shrinkage - Not applicable (e.g., for THETAs)
- `NA` - Could not be calculated

---

## 💡 Tips and Best Practices

### 1. Always Use YAML Specification

```{r tip1}
# ✅ Good - formatted output
spec_pk <- yaml::read_yaml(here::here("data", "spec", "pk-extend.yml"))
pk_res <- model_report(model_names = "run28", spec_pk = spec_pk)

# ❌ Less ideal - raw parameter names
pk_res <- model_report(model_names = "run28")
```

### 2. Check Models Exist First

```{r tip2}
# Check which models exist
available_models <- list.files("models", pattern = "^run.*\\.yaml$")
available_models <- gsub("\\.yaml$", "", available_models)

# Only report on existing models
pk_res <- model_report(
  model_names = available_models,
  spec_pk = spec_pk
)
```

### 3. Compare Similar Models

```{r tip3}
# Good - compare base model with covariate models
pk_res <- model_report(
  model_names = c("run10", "run11", "run12"),  # Base, +WT, +AGE
  spec_pk = spec_pk
)

# Less useful - compare very different models
# (different structures, different parameters)
```

### 4. Save Intermediate Results

```{r tip4}
# Save for later use
pk_res <- model_report(
  model_names = c("run85", "run90"),
  spec_pk = spec_pk
)

# Save as RDS
saveRDS(pk_res, "output/parameter_table.rds")

# Load later
pk_res <- readRDS("output/parameter_table.rds")
```

---

## 📚 See Also

- [YAML Specification Files](yaml-specification-files.html) - How to create pk-extend.yml
- [Complete Workflow](complete-workflow.html) - Full SCM example
- [Troubleshooting](troubleshooting.html) - Common issues and solutions
