---
title: "Understanding the Recovery System"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Understanding the Recovery System

CovariateSearcher includes an intelligent recovery system that automatically handles model estimation failures.

## How Recovery Works

### Detection Phase

The system monitors for several failure indicators:

1. **Infinite OFV**: OFV > 10^10
2. **Singular Matrix**: "SINGULAR CONVERGENCE CRITERIA"
3. **Error 134**: Numerical errors
4. **Minimization Failure**: "MINIMIZATION NOT SUCCESSFUL"

### Automatic Retry

When a failure is detected, the system:

```r
# 1. Marks original model as failed
model$status <- "failed"

# 2. Creates retry model with adjusted parameters
  original_model = run11
  retry_number = Creates run11001


# 3. Adjusts THETA initial estimate
# Original: THETA(1) = 0.1
# Retry:    THETA(1) = -0.1

# 4. Submits retry automatically if auto_retry = TRUE
```

## Example Recovery Workflow

### Scenario: AGE on CL fails

```{r example}
# Step 1: Original model (run11) fails
search_state$search_database %>%
  filter(model_name == "run11")
#   model_name  status  estimation_issue
#   run11       failed  OFV > 10^10

# Step 2: Retry created automatically (run11001)
search_state$search_database %>%
  filter(model_name == "run11001")
#   model_name  status       phase
#   run11001    in_progress  retry

# Step 3: If retry succeeds
search_state$search_database %>%
  filter(model_name == "run11001")
#   model_name  status     ofv      delta_ofv
#   run11001    completed  1234.5   -5.2
```

## Retry Strategies

### THETA Sign Adjustment

The most common issue is poor initial estimates. Changing the sign often helps:

```
Original THETA:
$THETA
0.1  ; AGE on CL

Retry THETA:
$THETA
-0.1  ; AGE on CL
```

### When Retries Fail

If retry also fails, the covariate relationship may be:

1. **Too complex**: Try simpler relationship (linear instead of power)
2. **Collinear**: Covariate correlated with another in model
3. **Wrong functional form**: Try different relationship type

## Monitoring Recovery

### Check Recovery Status

```{r monitor-recovery}
# View all retry models
retries <- search_state$search_database %>%
  filter(phase == "retry") %>%
  arrange(parent_model, model_name)

print(retries)
```

### Success Rate

```{r success-rate}
# Calculate recovery success rate
recovery_stats <- search_state$search_database %>%
  filter(phase == "retry") %>%
  count(status) %>%
  mutate(rate = n / sum(n))

print(recovery_stats)
```

## Disabling Auto-Retry

If you prefer manual control:

```{r manual}
search_state <- initialize_covariate_search(
  base_model_path = "run1",
  data_file_path = "data/derived/data.csv",
  covariate_search_path = "data/derived/covariates.csv",
  lookup_file = "data/spec/lookup.yaml"  # Optional: covariate labels
)

# Then manually create retries when needed
retry_model <- create_retry_model(
  search_state = search_state,
  original_model_name = "run11"
  # issue_type defaults to "estimation_error"
)

# Or disable automatic retry in the search function
search_result <- run_automated_scm_testing(
  search_state = search_state,
  auto_retry = FALSE  # Disable automatic retry for manual control
)
```

## Advanced: Custom Recovery

For complex cases, you can implement custom recovery:

```{r custom}
# 1. Identify problematic models
failed <- search_state$search_database %>%
  filter(status == "failed", phase != "retry")

# 2. For each failure, create custom retry
for (model in failed$model_name) {
  # Read original model
  model_path <- file.path(search_state$models_folder, model)
  
  # Custom modifications beyond THETA sign
  # - Adjust initial estimates
  # - Change estimation method
  # - Simplify covariate relationship
  
  # Create modified model
  # (custom code here)
}
```

## Best Practices

1. **Let auto-retry work first**: Usually fixes 60-80% of failures
2. **Check .lst files**: Understand why models failed
3. **Review base model**: Ensure it's stable before SCM
4. **Monitor RSE**: High RSE indicates estimation issues
5. **Use appropriate thresholds**: ΔOFV=3.84, RSE<50% are reasonable

## Recovery Statistics

Track recovery performance across your search:

```{r stats}
# Summary of all models
summary_stats <- search_state$search_database %>%
  group_by(phase, status) %>%
  summarize(
    count = n(),
    .groups = "drop"
  )

print(summary_stats)

# Successful recovery rate
successful_retries <- search_state$search_database %>%
  filter(phase == "retry", status == "completed") %>%
  nrow()

total_retries <- search_state$search_database %>%
  filter(phase == "retry") %>%
  nrow()

cat("Recovery success rate:", 
    round(100 * successful_retries / total_retries, 1), "%\n")
```

## Resuming After Recovery

When your search encounters failures and retries, you can pause and resume at any time:

### Pause and Resume Workflow

```{r resume-recovery}
# 1. Load existing search state (continue from last session)
search_state <- load_existing_search(
  base_model_path = "run1",
  data_file_path = "data/derived/data.csv",
  covariate_search_path = "data/derived/covariates.csv",
  models_folder = "models",
  lookup_file = "data/spec/lookup.yaml"  # Optional: restore covariate labels
)

# 2. Check status of recovery
retries_completed <- search_state$search_database %>%
  filter(phase == "retry", status == "completed") %>%
  arrange(desc(model_name))

print(retries_completed)

# 3. Continue search from last checkpoint
search_result <- run_automated_scm_testing(
  search_state = search_state,
  full_scm = TRUE,           # Continue with full SCM workflow
  starting_phase = "forward" # Start from forward selection
)
```

### Handling Persistent Failures

If the same covariate repeatedly fails across retries:

```{r persistent-failure}
# 1. Load search state
search_state <- load_existing_search(
  base_model_path = "run1",
  data_file_path = "data/derived/data.csv",
  covariate_search_path = "data/derived/covariates.csv",
  models_folder = "models"
)

# 2. Identify problematic covariate
persistent_failures <- search_state$search_database %>%
  filter(status == "failed") %>%
  group_by(covariate) %>%
  summarize(failure_count = n()) %>%
  arrange(desc(failure_count))

print(persistent_failures)

# 3. Remove problematic covariate from remaining search
search_state$search_config$excluded_covariates <- c(
  search_state$search_config$excluded_covariates,
  "AGE"  # Example: exclude AGE if it keeps failing
)

# 4. Continue with updated configuration
search_result <- run_automated_scm_testing(
  search_state = search_state,
  full_scm = TRUE
)
```

## FAQ: Resuming from Unfinished Steps

### Do I need to delete existing models when resuming?

**No.** When you call `load_existing_search()`, the system automatically:

1. **Detects existing models** in the models folder
2. **Loads the database** of all completed and failed models
3. **Updates the model counter** to the highest existing model number
4. **Continues with the next available number** for new models

Example:
```{r resume-no-cleanup}
# Session 1: Stop after step 4, models run1 through run25 exist
# (run1 = base, run2-run25 = models from steps 1-4)

# Session 2: Resume the next day
search_state <- load_existing_search(
  base_model_path = "run1",
  data_file_path = "data/derived/data.csv",
  covariate_search_path = "data/derived/covariates.csv",
  models_folder = "models"  # Contains run1-run25
)

# Model counter automatically set to 25
# Next model created will be run26 (start of step 5)
# No duplicates of run2-run25 will be created
```

### Will existing models be re-tested?

**No.** The system checks the database to see which models have already been tested:

- Completed models (with OFV values) are skipped
- Failed models are recorded but not re-created
- Only new covariates or covariate combinations are tested in new models

### Can I do a fresh start from a specific point?

If you want to restart a specific phase (e.g., re-run backward elimination), you can:

```{r fresh-start-phase}
# Option 1: Delete specific models from database before resuming
# (requires direct database manipulation - use with caution)

# Option 2: Create a new search with a different base model
search_state <- initialize_covariate_search(
  base_model_path = "run20",  # Start from run20 instead
  data_file_path = "data/derived/data.csv",
  covariate_search_path = "data/derived/covariates.csv",
  models_folder = "models_phase4_retry"  # Separate folder
)
```

