---
title: "Understanding the Recovery System"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Understanding the Recovery System

CovariateSearcher includes an intelligent recovery system that automatically handles model estimation failures.

## How Recovery Works

### Detection Phase

The system monitors for several failure indicators:

1. **Infinite OFV**: OFV > 10^10
2. **Singular Matrix**: "SINGULAR CONVERGENCE CRITERIA"
3. **Error 134**: Numerical errors
4. **Minimization Failure**: "MINIMIZATION NOT SUCCESSFUL"

### Automatic Retry

When a failure is detected, the system:

```r
# 1. Marks original model as failed
model$status <- "failed"

# 2. Creates retry model with adjusted parameters
retry_model <- create_retry_model(
  original_model = "run11",
  retry_number = 1  # Creates run11001
)

# 3. Adjusts THETA initial estimate
# Original: THETA(1) = 0.1
# Retry:    THETA(1) = -0.1

# 4. Submits retry automatically if auto_retry = TRUE
```

## Example Recovery Workflow

### Scenario: AGE on CL fails

```{r example}
# Step 1: Original model (run11) fails
search_state$search_database %>%
  filter(model_name == "run11")
#   model_name  status  estimation_issue
#   run11       failed  OFV > 10^10

# Step 2: Retry created automatically (run11001)
search_state$search_database %>%
  filter(model_name == "run11001")
#   model_name  status       phase
#   run11001    in_progress  retry

# Step 3: If retry succeeds
search_state$search_database %>%
  filter(model_name == "run11001")
#   model_name  status     ofv      delta_ofv
#   run11001    completed  1234.5   -5.2
```

## Retry Strategies

### THETA Sign Adjustment

The most common issue is poor initial estimates. Changing the sign often helps:

```
Original THETA:
$THETA
0.1  ; AGE on CL

Retry THETA:
$THETA
-0.1  ; AGE on CL
```

### When Retries Fail

If retry also fails, the covariate relationship may be:

1. **Too complex**: Try simpler relationship (linear instead of power)
2. **Collinear**: Covariate correlated with another in model
3. **Wrong functional form**: Try different relationship type

## Monitoring Recovery

### Check Recovery Status

```{r monitor-recovery}
# View all retry models
retries <- search_state$search_database %>%
  filter(phase == "retry") %>%
  arrange(parent_model, model_name)

print(retries)
```

### Success Rate

```{r success-rate}
# Calculate recovery success rate
recovery_stats <- search_state$search_database %>%
  filter(phase == "retry") %>%
  count(status) %>%
  mutate(rate = n / sum(n))

print(recovery_stats)
```

## Disabling Auto-Retry

If you prefer manual control:

```{r manual}
search_state <- initialize_new_search(
  base_model_path = "models/run1",
  data_file_path = "data/data.csv",
  covariate_search_path = "data/covariates.csv",
  auto_retry = FALSE  # Disable automatic retry
)

# Then manually create retries when needed
retry_model <- create_retry_model(
  search_state = search_state,
  model_name = "run11",
  retry_number = 1
)
```

## Advanced: Custom Recovery

For complex cases, you can implement custom recovery:

```{r custom}
# 1. Identify problematic models
failed <- search_state$search_database %>%
  filter(status == "failed", phase != "retry")

# 2. For each failure, create custom retry
for (model in failed$model_name) {
  # Read original model
  model_path <- file.path(search_state$models_folder, model)
  
  # Custom modifications beyond THETA sign
  # - Adjust initial estimates
  # - Change estimation method
  # - Simplify covariate relationship
  
  # Create modified model
  # (custom code here)
}
```

## Best Practices

1. **Let auto-retry work first**: Usually fixes 60-80% of failures
2. **Check .lst files**: Understand why models failed
3. **Review base model**: Ensure it's stable before SCM
4. **Monitor RSE**: High RSE indicates estimation issues
5. **Use appropriate thresholds**: ΔOFV=3.84, RSE<50% are reasonable

## Recovery Statistics

Track recovery performance across your search:

```{r stats}
# Summary of all models
summary_stats <- search_state$search_database %>%
  group_by(phase, status) %>%
  summarize(
    count = n(),
    .groups = "drop"
  )

print(summary_stats)

# Successful recovery rate
successful_retries <- search_state$search_database %>%
  filter(phase == "retry", status == "completed") %>%
  nrow()

total_retries <- search_state$search_database %>%
  filter(phase == "retry") %>%
  nrow()

cat("Recovery success rate:", 
    round(100 * successful_retries / total_retries, 1), "%\n")
```

