---
title: "YAML Specification Files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{YAML Specification Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# YAML Specification Files

CovariateSearcher reads variable and parameter metadata from YAML files shaped
like [yspec](https://metrumresearchgroup.github.io/yspec/) specifications. They
are what turns a bare `CL` into `CL/F (L/h)` with a description and a section,
and `SEXN level 2` into `SEXN Female`.

Three files are involved, and only two of them are yours to write:

| file | written by | read by |
|---|---|---|
| `lookup.yaml` | you | `build_covariate_reference_table()`, `model_add_cov()`, `model_report(lookup =)`, forest plots, boxplots |
| `pk-extend.yml` | you | `model_report(spec_pk =)`, forest and exposure summary tables |
| `tags.yaml` | the package | the search itself, to track which model carries which covariate |

Nothing here is required to *run* a search. A search with no specs works; its
output tables just carry raw NONMEM names.

---

## lookup.yaml — variables

One entry per variable in your dataset. The package reads two fields from it:

| field | used for |
|---|---|
| `values` | the levels a categorical covariate may take |
| `decode` | the human-readable name of each level, positionally matched to `values` |

`short`, `label` and `unit` are read too, by the forest and boxplot paths, for
axis and column labels.

Other yspec fields are safe to keep but are not read by this package. In
particular it looks for `unit` only — the `unit.text` and `unit.tex` variants
are yspec's own rendering fields and are ignored here, so a variable that
carries only those will come out unitless.

```yaml
SETUP__:
  description: Formoterol Population PK Analysis
  sponsor: AstraZeneca
  projectnumber: PT010
AGE:
  short: Age
  type: numeric
  unit: years
WT:
  short: Body weight
  type: numeric
  unit: kg
RACE:
  short: Race
  values: [1, 2, 3, 9]
  decode: ["White", "Black or African American", "Asian", "Other"]
SEXN:
  short: Sex
  values: [1, 2]
  decode: ["Male", "Female"]
```

### Categorical levels are validated against your data

`build_covariate_reference_table(..., yaml_data = spec)` checks every
categorical covariate against this file, and the checks are not advisory:

- a covariate with `Category = "cat"` that has **no entry at all** stops the
  build;
- a level present in your baseline data but **missing from `values`** stops the
  build, naming the level;
- a level declared in `values` but **never observed** in the data is a warning,
  not an error — it is usually a spec written for a larger population.

So a categorical covariate cannot enter a search until its levels are declared
here. That is deliberate: the level list decides how many THETAs the effect
needs and what each one is called.

### `SETUP__` is not read by the package

The `SETUP__` block, including `flags:` with `contcov`/`catcov`, is yspec
project metadata. CovariateSearcher never reads it, and it does **not** decide
which covariates get tested — that comes solely from the covariate table you
build with `build_covariate_reference_table()`.

It is still worth keeping, because your own code can use it. The boxplot
examples pull flags through yspec to decode a dataset in one call:

```{r flags}
spec  <- yspec::ys_load("data/spec/lookup.yaml")
flags <- yspec::pull_meta(spec, "flags")

dat <- decode_dataset(dat, spec, c(flags$catcov))
```

---

## pk-extend.yml — parameters

One entry per model parameter, controlling how `model_report()` renders it.

| field | effect on the table |
|---|---|
| `short` | the displayed name; combined with `unit` as `short (unit)` |
| `label` | the description column |
| `unit` | appended to `short` in parentheses |
| `comment` | which section the parameter is grouped under |

```yaml
SETUP__:
  max_nchar_label: 10000
KA:
  short: Ka
  label: Absorption rate constant
  unit: 1/hr
  comment: Typical parameters
CL:
  short: CL/F
  label: Apparent clearance
  unit: L/h
  comment: Typical parameters
V1:
  short: Vc/F
  label: Apparent central volume
  unit: L
  comment: Typical parameters
IIV_CL:
  short: CL/F CV%
  label: Inter-individual variability on clearance
  comment: Inter-individual variability
RUV_PROP:
  short: prop error
  label: Proportional residual error
  comment: Residual variability
```

### Exposure metrics live here too

The forest and exposure summary tables take quantity labels from a spec of the
same shape, so derived quantities belong in this file alongside the model
parameters. They need no `comment:` — they are never sectioned into a parameter
table:

```yaml
AUCss:
  short: AUC,ss
  label: steady-state area under the concentration-time curve
  unit: mg*hr/L
CMAXss:
  short: Cmax,ss
  label: steady-state maximum concentration
  unit: ng/mL
CMINss:
  short: Cmin,ss
  label: steady-state minimum concentration
  unit: ng/mL
```

The key must match the column name carrying that metric. See
[Forest Plots and Simulations](forest-plots.html).

### The keys must match your control stream

This is the one thing that silently produces a bad table. `model_report()` does
not read parameter names from the `.ext` file — it reads them from the
annotations in your control stream, taking the **second semicolon-separated
field** of each `$THETA`, `$OMEGA` and `$SIGMA` line:

```
$THETA
0.5  ; CL ; L/h ; LOG
2.0  ; V1 ; L   ; RATIO
```

Those names — `CL`, `V1` — are the keys `pk-extend.yml` must use. A key that
matches nothing in the control stream is ignored, and a parameter with no
matching key is reported back to you as "Parameters not found in spec_pk" and
falls through to its raw name.

The fourth field (`LOG`, `RATIO`) is separate: it drives back-transformation in
both `model_report()` and `plot_nonmem_iterations(transform = TRUE)`. See
[Model Reports and Diagnostics](model-reports-and-diagnostics.html).

### `comment` must be one of five exact strings

```
Typical parameters
Inter-individual variability
Correlation of random effects
Parameter-Covariate relationships
Residual variability
```

These are the section headers, and they appear in that order. Anything else —
"Random effects", "Residual error" — leaves the parameter without a section.
Covariate effects (`beta_*`) are assigned to "Parameter-Covariate
relationships" automatically and need no entry of their own; what they *do*
need is an entry for the **parameter** they act on, since the covariate row is
labelled from that parameter's `short` and `label`.

Omitting `comment` does **not** fall back to something sensible. A parameter
that appears in `spec_pk` without a `comment:` is left unsectioned and sinks to
the bottom of the table next to `OFV`. The name-based fallback — `THETA` to
typical parameters, `OMEGA` to IIV, `SIGMA` to residual variability — applies
only when no `spec_pk` is supplied at all. Once you pass a spec, every
parameter in it needs its `comment:`.

---

## tags.yaml — generated, not written

`initialize_covariate_search()` and `add_covariates_to_search()` regenerate
`data/spec/tags.yaml` from the covariate search table via
`generate_tags_from_covariate_search()`. It maps each `beta_` tag to its
covariate-parameter pair and is how the search knows which covariates a given
model carries.

```yaml
# Tags configuration file
# Auto-generated from covariate search table

## Covariates
beta_WT_CL: "WT_CL"                # Continuous, power, time-independent
beta_SEXN_CL: "SEXN_CL"            # Categorical 2-level, linear, time-independent
beta_WT_V1: "WT_V1"                # Continuous, power, time-independent
```

Do not hand-edit it. Change the covariate table and let it be rewritten —
editing the file directly desynchronises it from the table that drives the
search.

---

## Checking your files

```{r validate}
lookup  <- yaml::read_yaml("data/spec/lookup.yaml")
spec_pk <- yaml::read_yaml("data/spec/pk-extend.yml")

# Do the parameter keys match the control stream annotations?
names(spec_pk)

# Are all five section names spelled exactly?
unique(unlist(lapply(spec_pk, function(x) x$comment)))
```

Both files are passed as **objects**, never as paths:

```{r pass_objects}
pk_res <- model_report(
  model_names = "run28",
  spec_pk     = spec_pk,
  lookup      = lookup
)
```

---

## See also

- [Model Reports and Diagnostics](model-reports-and-diagnostics.html) — what
  `spec_pk` and `lookup` do to the output
- [Complete Workflow](complete-workflow.html) — where the specs enter a search
- [Forest Plots and Simulations](forest-plots.html) — the other consumer of
  `pk-extend.yml`
- [yspec documentation](https://metrumresearchgroup.github.io/yspec/) and the
  [yspec book](https://metrumresearchgroup.github.io/ysp-book/)
