---
title: "YAML Specification Files"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# YAML Specification Files for CovariateSearcher

CovariateSearcher uses YAML specification files from the [yspec](https://metrumresearchgroup.github.io/yspec/) package (developed by Metrum Research Group) to enhance documentation and create formatted output tables.

## 📋 Overview

The YAML specification system provides:

- ✅ **Standardized variable definitions** with units and descriptions
- ✅ **Formatted output tables** for parameters and results
- ✅ **Categorical variable decoding** (e.g., 1 = "White", 2 = "Black")
- ✅ **Automatic documentation** generation

---

## 🗂️ Required Files

You need three YAML files in your `data/spec/` directory:

```
data/spec/
├── lookup.yml        # Variable definitions and decoding
├── pk-extend.yml     # Parameter formatting for output tables
└── pk.yml            # Main specification file
```

---

## 📄 File 1: lookup.yml

Defines variables, their types, units, and categorical value decoding.

### Example Structure

```yaml
SETUP__:
  description: yspec internal column database for PT010
NUM:
  short: Row number
  type: numeric
OPROJID:
  short: Project name
  type: character
RACE:
  short: Race
  values: [1, 2, 3, 9]
  decode: ["White", "Black or African American", "Asian", "Other"]
EGFR:
  short: Estimated GFR
  type: numeric
  unit.text: "mL/min/1.73m^2"
BBSA:
  short: Body surface area
  unit: m2
  unit.tex: "m^2"
```


### Common Variables to Include

**Demographics:**
```yaml
AGE:
  short: Age
  type: numeric
  unit: years
WT:
  short: Body weight
  type: numeric
  unit: kg
SEX:
  short: Sex
  values: [0, 1]
  decode: ["Female", "Male"]
```

**Clinical Covariates:**
```yaml
CRCL:
  short: Creatinine clearance
  type: numeric
  unit: mL/min
BMI:
  short: Body mass index
  type: numeric
  unit.text: "kg/m^2"
```

---

## 📄 File 2: pk-extend.yml

Defines parameter formatting for output tables. This creates nicely formatted parameter names with units and descriptions.

### Example Structure

```yaml
SETUP__:
  max_nchar_label: 10000
DVNORM:
  short: Dose-normalized concentration
  unit: ng/mL/mg
AUCss:
  short: AUC,ss
  label: steady-state area under the concentration-time curve
  unit: mg*hr/L
CMAXss:
  short: Cmax,ss
  label: steady-state maximum concentration
  unit: ng/mL
CMINss:
  short: Cmin,ss
  label: steady-state minimum concentration
  unit: ng/mL
KA:
  short: Ka
  label: Absorption rate constant
  unit: 1/hr
  comment: Typical parameters
Frel:
  short: Frel
  label: Relative bioavailability
  unit: percent
  comment: Typical parameters
CL:
  short: CL/F
  label: Apparent clearance
  unit: L/h
  comment: Typical parameters
V1:
  short: Vc/F
  label: Apparent central volume
  unit: L
  comment: Typical parameters
V2:
  short: Vp1/F
  label: Apparent peripheral volume
  unit: L
  comment: Typical parameters
```

### Key Fields for Parameters

| Field     | Description         | Example              |
|-----------|---------------------|----------------------|
| `short`   | Abbreviated name    | "CL/F"               |
| `label`   | Full description    | "Apparent clearance" |
| `unit`    | Unit of measurement | "L/h"                |
| `comment` | Category or note    | "Typical parameters" |

comments should be selected from this list:
Typical parameters"
Inter-individual variability
Correlation of random effects
Parameter-Covariate relationships
Residual variability

### Common Parameters to Include


```yaml
CL:
  short: CL/F
  label: Apparent clearance
  unit: L/h
  comment: Typical parameters
V:
  short: V/F
  label: Apparent volume of distribution
  unit: L
  comment: Typical parameters
KA:
  short: Ka
  label: Absorption rate constant
  unit: 1/hr
  comment: Typical parameters
IIV_CL:
  short:CL/F CV%
  label: Inter-individual variability on clearance
  comment: Random effects
IIV_V:
  short: V/F CV%
  label: Inter-individual variability on volume
  comment: Random effects
RUV_PROP:
  short: Prop error
  label: Proportional residual error
  comment: Residual error
```

---

## 📄 File 3: pk.yml

Main specification file that ties everything together and defines variable categories for plots and tables.

### Example Structure

```yaml
SETUP__:
  lookup_file: [lookup.yml]
  description: Formoterol Population PK Analysis
  sponsor: AstraZeneca
  projectnumber: PT010
  extend_file: pk-extend.yml
  flags:
    contcov: [AGE, SCR, CRCL, EGFR, BEGFR, BSCR, BCRCL, BWT, BBMI, BBSA]
    catcov: [SEXM, RACE, ETHNIC, SMOKING, DOSE, FORM, TRTID, ICS, ASTHMA]
    diagContCov: [AGE, BEGFR, BSCR, BCRCL, BWT, BBMI, BBSA]
    diagCatCov: [SEXM, RACE, ETHNIC, SMOKING, DOSE, FORM, TRTID, ICS, ASTHMA]
NUM:
OPROJID:
OSTUDYID:
STUDYID:
USID:
ID:
IND:
AGE:
SEXM:
RACE:
ETHNIC:
SMOKING:
```

### SETUP__ Section

| Field           | Description                          |
|-----------------|--------------------------------------|
| `lookup_file`   | lookup.yml                           |
| `description`   | Project description                  |
| `sponsor`       | Sponsor name                         |
| `projectnumber` | Project identifier                   |
| `extend_file`   | pk-extend.yml                        |
| `flags`         | Variable categories for plots/tables |

### Flags Categories

**`contcov`**: Continuous covariates for covariate modeling
```yaml
contcov: [AGE, WT, BMI, CRCL, EGFR]
```

**`catcov`**: Categorical covariates for covariate modeling
```yaml
catcov: [SEX, RACE, ETHNIC, SMOKING, DOSE]
```

**`diagContCov`**: Continuous covariates for diagnostic plots
```yaml
diagContCov: [AGE, WT, BMI, CRCL]
```

**`diagCatCov`**: Categorical covariates for diagnostic plots
```yaml
diagCatCov: [SEX, RACE, SMOKING, DOSE]
```



---

## 🎯 How CovariateSearcher Uses YAML Files

### 1. Variable Decoding
Categorical variables are automatically decoded in output tables:

```
Instead of: RACE = 1, 2, 3
Shows:      RACE = White, Black or African American, Asian
```

### 2. Parameter Formatting
Output tables show formatted parameter names:

```
Instead of: CL = 5.2 (units unclear)
Shows:      CL/F [L/h] = 5.2 (Apparent clearance)
```

### 3. Covariate Categorization
Automatically identifies which covariates to test:

```yaml
flags:
  contcov: [AGE, WT, BMI]  # Test these continuous covariates
  catcov: [SEX, RACE]      # Test these categorical covariates
```

---

## 📝 Creating Your YAML Files

### Step 1: Start with lookup.yml

List all variables in your dataset:

```yaml
SETUP__:
  description: My PK Analysis
ID:
  short: Subject ID
  type: numeric
TIME:
  short: Time
  type: numeric
  unit: hr
DV:
  short: Concentration
  type: numeric
  unit: ng/mL
AGE:
  short: Age
  type: numeric
  unit: years
WT:
  short: Weight
  type: numeric
  unit: kg
SEX:
  short: Sex
  values: [0, 1]
  decode: ["Female", "Male"]
```

### Step 2: Create pk-extend.yml

Define your model parameters:

```yaml
SETUP__:
  max_nchar_label: 10000
CL:
  short: CL/F
  label: Apparent clearance
  unit: L/h
V:
  short: V/F
  label: Apparent volume
  unit: L
IIV_CL:
  short: CL CV%
  label: IIV on clearance
RUV_PROP:
  short: prop error
  label: Proportional error
```

### Step 3: Create pk.yml

Tie everything together:

```yaml
SETUP__:
  lookup_file: [lookup.yml]
  description: My Population PK Analysis
  extend_file: pk-extend.yml
  flags:
    contcov: [AGE, WT, BMI, CRCL]
    catcov: [SEX, RACE, SMOKING]
ID:
TIME:
DV:
AGE:
WT:
SEX:
RACE:
```

---

## ✅ Best Practices

### 1. Match Parameter Names
Ensure names in `pk-extend.yml` match your NONMEM model:

```yaml
# If your model has:
$THETA
0.5 ; CL ; L/h ; LOG

# Your pk-extend.yml should have:
CL:
  short: CL/F
  label: Apparent clearance
  unit: L/h
```

### 2. Include All Covariates
List all potential covariates in both `lookup.yml` and `pk.yml`:

```yaml
# In lookup.yml - define them
AGE:
  short: Age
  type: numeric
  unit: years

# In pk.yml - reference them
AGE:
```

### 3. Use Consistent Units
Match units across files:

```yaml
# lookup.yml
WT:
  unit: kg

# pk-extend.yml
CL:
  unit: L/h  # Not L/hr or L/hour
```

### 4. Decode All Categorical Variables
Provide meaningful labels:

```yaml
RACE:
  values: [1, 2, 3, 9]
  decode: ["White", "Black or African American", "Asian", "Other"]
  # NOT: ["1", "2", "3", "9"]
```

---

## 🔍 Validation

Check your YAML files are valid:

```{r validate, eval=FALSE}
library(yspec)

# Load and validate
spec <- ys_load("data/spec/pk.yml")

# Check for issues
ys_check(spec)

# View summary
summary(spec)
```

---

## 📚 Additional Resources

- **yspec Documentation**: https://metrumresearchgroup.github.io/yspec/
- **yspec Book**: https://metrumresearchgroup.github.io/ysp-book/
- **Examples**: See the yspec package for more examples

---

## 🎯 Summary

**Three files needed:**

1. **lookup.yml** - Variable definitions, types, units, decoding
2. **pk-extend.yml** - Parameter formatting for output tables
3. **pk.yml** - Main spec file that references the others

**Key benefits:**

- Automatic variable decoding
- Formatted output tables
- Consistent documentation
- Enhanced reporting

**Location:** `data/spec/` directory in your project

---

## 💡 Quick Start Template

Minimal working set:

**lookup.yml:**
```yaml
SETUP__:
  description: My Analysis
AGE:
  short: Age
  type: numeric
  unit: years
WT:
  short: Weight
  type: numeric
  unit: kg
```

**pk-extend.yml:**
```yaml
SETUP__:
  max_nchar_label: 10000
CL:
  short: CL/F
  label: Apparent clearance
  unit: L/h
```

**pk.yml:**
```yaml
SETUP__:
  lookup_file: [lookup.yml]
  extend_file: pk-extend.yml
  flags:
    contcov: [AGE, WT]
AGE:
WT:
```

Build from this template as your analysis grows!
